Generating Images From Captions\\ With Attention

### 2015-10-06 iclr-paper.tex
update model sec

\subsection{Language Model: the Bidirectional Attention RNN}

The input caption sentences are fed into a deterministic Bidirectional LSTM\cite{} that encodes the variable size sentences into the vector representation $s$. Bidirectional LSTM consists of one Forward LSTM and Backward LSTM which combine information from past and future respectively. The Forward LSTM computes the sequence of forward hidden states $[\overrightarrow{h}^{lang}_{1}, \overrightarrow{h}^{lang}_{2}, ..., \overrightarrow{h}^{lang}_{N}]$ , whereas the Backward LSTM computes the sequence of backward hidden states $[\overleftarrow{h}^{lang}_{1}, \overleftarrow{h}^{lang}_{2}, ..., \overleftarrow{h}^{lang}_{N}]$. Then these hidden states are concatenated together into the sequence $[\hlang_{1}, \hlang_{2}, ..., \hlang_{N}]$, where $\hlang_{n} = [\overrightarrow{h}^{lang}_{n}, \overleftarrow{h}^{lang}_{n}], 1\leq n\leq N$.

The final representation of the sentence is calculated as follows:
\begin{align}
s_{t} = \alpha_{1}\hlang_{1} + \alpha_{2}\hlang_{2} + ... + \alpha_{N}\hlang_{N}
\end{align}
where $\hlang_{0}$ is initialized to the learned bias.
Setting $\alpha_{1...N}$ to $\frac{1}{N}$ turns the encoder into the vanilla model introduced in \citep{cho_mt} without the attention. We will describe how $\alpha_{1...N}$ are calculated in the next section.

@@

The image generative model is a DRAW model, which is a stochastic recurrent neural network that consists of Inference LSTM that infers the distribution of latent variables of image $x$ given $y$ and then the Generator LSTM that uses the inferred latent variables in order to reconstruct the image $x$ given $y$. We extend the DRAW network to include the $align$ function by \cite{bahdanau_mt} which used to compute the alignment between the input caption and intermediate image generative steps:

Formally, the decoder iteratively computes the following equations for $t=1,...,T$
\begin{align}
\label{eq:x_hat}
\hat{x}_t &= x-\sigmoid(\canv_{t-1})\\
\label{eq:read}
r_t &= \readop(x_t, \hat{x}_t, \hdec_{t-1})\\
\henc_t &= \encoder(\henc_{t-1}, [r_t, \hdec_{t-1}])\\
\lat_t &\sim \post(\Lat_t|\henc_t)\\
\hdec_t &= \decoder(\hdec_{t-1}, z_t, s_{t-1})\\
s_{t} &= align(\hdec_{t-1}, \hlangall) \\
\label{eq:write}
\canv_t &= \canv_{t-1} + \writeop(\hdec_t)
\end{align}

where $\readop$ and $\writeop$ are the same attention operators as in \citep{gregor_draw}, $\hlangall = [\hlang_{1}, \hlang_{2}, ..., \hlang_{n}]$ and $\canv_{0}, \hdec_{0}, \henc_{0}$ are initialized to the learned biases.

The cooresponding alignment probabilities $\alpha_{1...n}$ at each step are obtained by:
\begin{align}
e_{tj} &= v^{T}tanh(U\hlang_{j} + W\hdec_{t} + b)\\
\alpha_{j} &= \frac{exp(e_{tj})}{\sum_{j=1}^{N}exp(e_{tj})}.
\end{align}

@@

The latent loss is a negative sum of Kullback--Leibler divergence terms between distribution $\post(\Lat_t|\henc_t)$ and some prior distribution ${\prior(\Lat_t)}$ over time $t=1,...,T$, which can be seen as a regularization term. Since the patches drawn on canvas over time are not independent of each other, naturally the sufficient statistics of the prior distribution at time $t$ should be dependent on the sufficient statistics of the prior distribution at time $t-1$. Therefore, instead of setting $\prior(\Lat_1), ..., \prior(\Lat_T)$ to be independent unit gaussian distributions, the mean and variance of $\prior(\Lat_t)$ depends on the $\hdec_{t-1}$, which forms a Markov chain $\prior(\Lat_1), \prior(\Lat_2\given\Lat_1), ..., \prior(\Lat_T\given\Lat_{T-1})$ as in \citep{bachman_sdm}, where 
\begin{align}
\mu_{t}^{prior} &= tanh(W_{\mu}\hdec_{t-1})\\
\sigma_{t}^{prior} &= exp(tanh(W_{\sigma}\hdec_{t-1})) 
\end{align}
Overall, the loss function $\loss$ is calculated as follows:
\begin{align}
\loss =  &\expectation_{Q(\Lat_{1:t}\given\icaption,\oimage)}\left[ - \log p(\oimage\given\icaption,\Lat_{1:T}) + \sum_{t=2}^T\kldiv\left(\post(\Lat_t\given\Lat_{t-1},\icaption,\oimage)\klBars\prior(\Lat_t\given\Lat_{t-1},\icaption)\right) \right] \nonumber \\& + \kldiv\left(\post(\Lat_1\given\icaption,\oimage)\klBars\prior(\Lat_1\given\icaption)\right).
\end{align}
The expectation can be approximate by $\numSamples$ Monte Carlo samples $\LatSample_{1:T}$ from $\post(\Lat_{1:T}\given\icaption, \oimage)$:
\begin{align}
\loss \approx  &\frac{1}{\numSamples}\sum_{\sampleIdx=1}^\numSamples\left[ - \log p(\oimage\given\icaption,\LatSample^\sampleIdx_{1:T}) + \sum_{t=2}^T\kldiv\left(\post(\Lat_t\given\LatSample^\sampleIdx_{t-1},\icaption,\oimage)\klBars\prior(\Lat_t\given\LatSample^\sampleIdx_{t-1},\icaption)\right) \right] \nonumber \\& + \kldiv\left(\post(\Lat_1\given\icaption,\oimage)\klBars\prior(\Lat_1\given\icaption)\right).
\end{align}
\comm{
\begin{align}
\loss &= -\sum_{t=1}^{T}D_{KL}(\post(\Lat_t|\henc_t,s_{t-1})\,||\,\prior(\Lat_t)) + \frac{1}{L}\sum_{l=1}^{L}log\,p(x_{t}|y,z)\\
&=
\frac{1}{2}\sum_{t=1}^{T}(1 - 2\,log\,\sigma_{t}^{prior} + 2\,log\,\sigma_{t} - \frac{exp(2\,log\,\sigma_{t}) + (\mu_{t} - \mu_{t}^{prior})^{2}}{exp(2\,log\,\sigma_{t}^{prior})}) + \frac{1}{L}\sum_{l=1}^{L}log\,p(x_{t}|y,z)
\end{align}
}

>>>>

\subsection{Language Model: the Bidirectional Attention RNN}
\label{sec:lang}
The input caption sentences are fed into a deterministic Bidirectional LSTM\cite{} that encodes the variable size sentences into the vector representation $s$. Bidirectional LSTM consists of one Forward LSTM and Backward LSTM which combine information from past and future respectively. The Forward LSTM computes the sequence of forward hidden states $[\overrightarrow{h}^{lang}_{1}, \overrightarrow{h}^{lang}_{2}, ..., \overrightarrow{h}^{lang}_{N}]$ , whereas the Backward LSTM computes the sequence of backward hidden states $[\overleftarrow{h}^{lang}_{1}, \overleftarrow{h}^{lang}_{2}, ..., \overleftarrow{h}^{lang}_{N}]$. Then these hidden states are concatenated together into the sequence $[\hlang_{1}, \hlang_{2}, ..., \hlang_{N}]$, where $\hlang_{n} = [\overrightarrow{h}^{lang}_{n}, \overleftarrow{h}^{lang}_{n}], 1\leq n\leq N$.

@@

The DRAW network \cite{gregor_draw} is a sequential probabilistic model generating images by accumulating the output at each iterative step. While the original DRAW network assumes the latent variables are independent, it has shown in \citep{bachman_sdm} the model performance is improved by including the dependencies of latent variables.

We extended the architecture the DRAW network generative process to include additional input caption from the language model described in Sec. (\ref{sec:lang}). Similarly to the original DRAW network, the conditional DRAW network is a stochastic recurrent neural network that consists of Inference LSTM that infers the distribution of latent variables of image $x$ given $y$ and then the Generator LSTM that uses the inferred latent variables in order to reconstruct the image $x$ given $y$. The $align$ function is used to compute the alignment between the input caption and intermediate image generative steps as in \cite{bahdanau_mt}:

Formally, the image is generated by iteratively computing the following equations for $t=1,...,T$
\begin{align}
\label{eq:x_hat}
\hat{x}_t &= x-\sigmoid(\canv_{t-1})\\
\label{eq:read}
r_t &= \readop(x_t, \hat{x}_t, \hdec_{t-1})\\
\henc_t &= \encoder(\henc_{t-1}, [r_t, \hdec_{t-1}])\\
\lat_t &\sim \post(\Lat_t|\henc_t)\\
\hdec_t &= \decoder(\hdec_{t-1}, z_t, s_{t-1})\\
s_{t} &= align(\hdec_{t-1}, \hlangall) \\
\label{eq:write}
\canv_t &= \canv_{t-1} + \writeop(\hdec_t)
\end{align}

where $\readop$ and $\writeop$ are the same attention operators as in \citep{gregor_draw}.
Given the caption representation from the language model, $\hlangall = [\hlang_{1}, \hlang_{2}, ..., \hlang_{N}]$, the $algin$ operator computes the final sentence representation $s_t$ through a weighted sum using alignment probabilities $\alpha_{1...N}$:
\begin{align}
s_t=align(\hdec_{t-1}, \hlangall) = \alpha_{1}\hlang_{1} + \alpha_{2}\hlang_{2} + ... + \alpha_{N}\hlang_{N}.
\end{align}

The corresponding alignment probabilities $\alpha_{1...n}$ at each step are obtained by:
\begin{align}
e_{tj} &= v^{T}tanh(U\hlang_{j} + W\hdec_{t} + b)\\
\alpha_{j} &= \frac{exp(e_{tj})}{\sum_{j=1}^{N}exp(e_{tj})}.
\end{align}

Here $\hlang_{0}$ is initialized to the learned bias.
Setting $\alpha_{1...N}$ to $\frac{1}{N}$ turns the encoder into the vanilla model introduced in \citep{cho_mt} without the attention.

@@

The latent loss $\lloss$ is a negative sum of Kullback--Leibler divergence terms between distribution $\post(\Lat_t|\henc_t)$ and some prior distribution ${\prior(\Lat_t)}$ over time $t=1,...,T$, which can be seen as a regularization term. Since the patches drawn on canvas over time are not independent of each other, naturally the sufficient statistics of the prior distribution at time $t$ should be dependent on the sufficient statistics of the prior distribution at time $t-1$. Therefore, instead of setting $\prior(\Lat_1), ..., \prior(\Lat_T)$ to be independent unit gaussian distributions, the mean and variance of $\prior(\Lat_t)$ depends on the $\hdec_{t-1}$, which forms a Markov chain $\prior(\Lat_1), \prior(\Lat_2\given\Lat_1), ..., \prior(\Lat_T\given\Lat_{T-1})$ as in \citep{bachman_sdm}, where 
\begin{align}
\mu_{t}^{prior} &= tanh(W_{\mu}\hdec_{t-1})\\
\sigma_{t}^{prior} &= exp(tanh(W_{\sigma}\hdec_{t-1})) 
\end{align}
Overall, the loss function $\loss$ is calculated as follows:
\begin{align}
\loss =  &\expectation_{Q(\Lat_{1:t}\given\icaption,\oimage)}\left[ - \log p(\oimage\given\icaption,\Lat_{1:T}) + \sum_{t=2}^T\kldiv\left(\post(\Lat_t\given\Lat_{t-1},\icaption,\oimage)\klBars\prior(\Lat_t\given\Lat_{t-1},\icaption)\right) \right] \nonumber \\& + \kldiv\left(\post(\Lat_1\given\icaption,\oimage)\klBars\prior(\Lat_1\given\icaption)\right).
\end{align}
The expectation can be approximate by $\numSamples$ Monte Carlo samples $\LatSample_{1:T}$ from $\post(\Lat_{1:T}\given\icaption, \oimage)$:
\begin{align}
\loss \approx  &\frac{1}{\numSamples}\sum_{\sampleIdx=1}^\numSamples\left[ - \log p(\oimage\given\icaption,\LatSample^\sampleIdx_{1:T}) + \sum_{t=2}^T\kldiv\left(\post(\Lat_t\given\LatSample^\sampleIdx_{t-1},\icaption,\oimage)\klBars\prior(\Lat_t\given\LatSample^\sampleIdx_{t-1},\icaption)\right) \right] \nonumber \\& + \kldiv\left(\post(\Lat_1\given\icaption,\oimage)\klBars\prior(\Lat_1\given\icaption)\right).
\end{align}
\comm{
\begin{align}
\loss &= -\sum_{t=1}^{T}D_{KL}(\post(\Lat_t|\henc_t,s_{t-1})\,||\,\prior(\Lat_t)) + \frac{1}{L}\sum_{l=1}^{L}log\,p(x_{t}|y,z)\\
&=
\frac{1}{2}\sum_{t=1}^{T}(1 - 2\,log\,\sigma_{t}^{prior} + 2\,log\,\sigma_{t} - \frac{exp(2\,log\,\sigma_{t}) + (\mu_{t} - \mu_{t}^{prior})^{2}}{exp(2\,log\,\sigma_{t}^{prior})}) + \frac{1}{L}\sum_{l=1}^{L}log\,p(x_{t}|y,z)
\end{align}
}
